---
title: "Storm Data Evaluation"
author: "Ron Vaughan"
date: "December 3, 2017"
output:
  md_document:
    variant: markdown_github
---
## 1: Synopsis
The goal is to explore the [NOAA Storm Database](https://www.ncdc.noaa.gov/stormevents/) and answer questions about severe weather events the effects of severe weather events on both population and economy. The database covers the time period between 1950 and November 2011.

### 1.1 Data Analysis

The following analysis investigates which types of severe weather events (EVTYPE variable) across the U.S. are most harmful to:

1. Population Health 
2. Greatest Economic Consequences


## 2: Data Processing

Source Data: [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)
National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
National Climatic Data Center Stsorm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

### 2.1: Data Loading

Download the raw data file and extract the data into a dataframe, then convert to a data.table
```{r}
library("data.table")
library("ggplot2")
library("plyr")
library("gridExtra")
library("grid")

fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(fileUrl, destfile = paste0(getwd(), '/repdata%2Fdata%2FStormData.csv.bz2'))
data <- read.csv(bzfile("repdata%2Fdata%2FStormData.csv.bz2"))
stormData <- as.data.table(data)
```

### 2.2: Data Subsetting

The analysis focuses only on the health and economic consequences of severe weather events so we subset to reduce the size of the dataset. 
```{r}
# Subsetting the data
stormDR <- stormData[,c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")]
```

### 2.3 Population Health

Fatalities and injuries are summarized according to event type, and ordered in decreasing value.
```{r}
harmToHealth <- ddply(stormDR, .(EVTYPE), summarize,fatalities = sum(FATALITIES),injuries = sum(INJURIES))
fatal <- harmToHealth[order(harmToHealth$fatalities, decreasing = T), ]
injury <- harmToHealth[order(harmToHealth$injuries, decreasing = T), ]
```

### 2.3: Economic Consequences

Exponential data is stored in separate columns with letter values, so we ,need to convert the letters to numbers.

```{r}
getExp <- function(e) {
    if (e %in% c("h", "H"))
        return(2)
    else if (e %in% c("k", "K"))
        return(3)
    else if (e %in% c("m", "M"))
        return(6)
    else if (e %in% c("b", "B"))
        return(9)
    else if (!is.na(as.numeric(e))) 
        return(as.numeric(e))
    else if (e %in% c("", "-", "?", "+"))
        return(0)
    else {
        stop("Invalid value.")
    }
}
```

Now property and crop damages are calculated.
```{r}
propExp <- sapply(stormDR$PROPDMGEXP, FUN=getExp)
stormDR$propDamage <- stormDR$PROPDMG * (10 ** propExp)
cropExp <- sapply(stormDR$CROPDMGEXP, FUN=getExp)
stormDR$cropDamage <- stormDR$CROPDMG * (10 ** cropExp)
```

We summarize the damages for crops and property by event type by excluding the events which did not cause finalcial impact.
```{r}
economicDamage <- ddply(stormDR, .(EVTYPE), summarize,propDamage = sum(propDamage), cropDamage = sum(cropDamage))
economicDamage <- economicDamage[(economicDamage$propDamage > 0 | economicDamage$cropDamage > 0), ]
```

Econommic data is sorted.
```{r}
propDmgSorted <- economicDamage[order(economicDamage$propDamage, decreasing = TRUE), ]
cropDmgSorted <- economicDamage[order(economicDamage$cropDamage, decreasing = TRUE), ]
```

## 3: Results

### 3.1: Events that are Most Harmful to Population Health

Calculating and showing top 5 total injuries and fatalities.
```{r}
head(injury[, c("EVTYPE", "injuries")],5)
head(fatal[, c("EVTYPE", "fatalities")],5)
```
It's apparent from the header results that TORNADOs generated the most injuries and fatalities.

Plotting the top 10 results for greater clarity.
```{r}
p1 <- ggplot(data=head(injury,10), aes(x=reorder(EVTYPE, injuries), y=injuries)) +
   geom_bar(fill="olivedrab",stat="identity")  + coord_flip() + 
    ylab("Total number of injuries") + xlab("Event type") +
    ggtitle("Health impact of weather events in the US - Top 10") +
    theme(legend.position="none")

p2 <- ggplot(data=head(fatal,10), aes(x=reorder(EVTYPE, fatalities), y=fatalities)) +
    geom_bar(fill="red4",stat="identity") + coord_flip() +
    ylab("Total number of fatalities") + xlab("Event type") +
    theme(legend.position="none")

grid.arrange(p1, p2, nrow =2)
```

### 3.2: Events that have the Greatest Economic Consequences

Calculating and showing the top 5 events for ecomomic damages.
```{r}
head(propDmgSorted[, c("EVTYPE", "propDamage")], 5)
head(cropDmgSorted[, c("EVTYPE", "cropDamage")], 5)
```

Plotting the top 10 results for greater clarity.
```{r}
p3 <- ggplot(data=head(propDmgSorted,10), aes(x=reorder(EVTYPE, propDamage), y=log10(propDamage), fill=propDamage )) +
    geom_bar(fill="darkred", stat="identity") + coord_flip() +
    xlab("Event type") + ylab("Property damage in dollars (log10)") +
    ggtitle("Economic impact of weather events in the US - Top 10") +
    theme(plot.title = element_text(hjust = 0))

p4 <- ggplot(data=head(cropDmgSorted,10), aes(x=reorder(EVTYPE, cropDamage), y=cropDamage, fill=cropDamage)) +
    geom_bar(fill="goldenrod", stat="identity") + coord_flip() + 
    xlab("Event type") + ylab("Crop damage in dollars") + 
    theme(legend.position="none")

grid.arrange(p3, p4, ncol=1, nrow =2)
```